---
title: "README"
author: "Isabelle Boulangeat"
date: "`r format(Sys.time(), '%d/%m/%y')`"
output:
  html_document:
    keep_md: yes
    variant: markdown_github
  pdf_document: default
editor_options:
  chunk_output_type: console
always_allow_html: yes
---
### projeter les obs dans l'espace bioclimatique

```{r setup, include=FALSE}
library(knitr)
# library(kableExtra)
knitr::opts_chunk$set(echo = TRUE)
# library(Jmisc)
library(tidyr)
sapply(list.files("R_fct"), function(x)source(paste0("R_fct/",x)))
require(sp)
require(dplyr)
require(ggplot2)
require(ggmap)
require(sf)
require(raster)
require(terra)
require(stars)
require(starsExtra)
require(shadow)
require(rgdal)
require(mapview)
```

TODO
- choisir projection UTM appropriée pour europe / transfrontalier
- add other bioclim vars ?? cf justification
- check if soil vars available
- projeter les obs dans l'espace climatique
  - check obs anciennes VS recentes
  - check mercantour VS savoie
  - check obs France VS Italy

MODEL
- variable selection??
- check effect precision (P=point, C=commune,  A? N? T?) id_precision
- add data italy
  - intégrer au modèle complet
  - projeter en France

## Chargement des données
```{r load}
pts = read.table("_data/export_478_24032021_151335.txt", sep='\t', h=TRUE)
sites_points = pts %>% dplyr::select(id_releve, date_releve_deb, lon_wgs84, lat_wgs84, x_l93, y_l93, id_precision)

sites_points1 = sites_points %>% filter(!is.na(lon_wgs84))
coordinates(sites_points1)= c("lon_wgs84","lat_wgs84")
proj4string(sites_points1) = CRS("+proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs")
#sites_points1 = spTransform(sites_points1, CRS("+init=epsg:4326"))

sites_df <- data.frame(sites_points1)
require(tmaptools)
bbox <- tmaptools::bb(xlim = c(5.68,  8.40), ylim = c(43.60, 46.12))

```


<!-- ## Répartition des observations -->
<!-- # ```{r obs,fig=TRUE} -->
<!-- # bbox_map <- c(left = 5.68,  right = 8.40, bottom = 43.60, top= 46.12) -->
<!-- # p <- ggmap(get_map(bbox_map, source = 'osm', -->
<!-- #                          maptype ='terrain', -->
<!-- #                          color = 'color')) -->
<!-- # p + geom_point(data=sites_df,  aes(x=lon_wgs84, y=lat_wgs84), size=1)  -->
<!-- # ``` -->


```{r leaflet, fig = TRUE}
library(leaflet) 
leaflet(sites_df)%>%
  addProviderTiles("OpenStreetMap.HOT")%>%
  setView(lng=5.5,lat=45,zoom=6) %>%
  addCircleMarkers(lng = ~lon_wgs84, lat = ~lat_wgs84, popup = ~date_releve_deb, radius = 1, opacity = 0.8, color = "red")
```

<!-- # Providers:  -->
<!-- #   OpenTopoMap -->
<!-- #   Thunderforest.Landscape -->
<!-- # Jawg.Terrain -->
<!-- #GeoportailFrance.plan -->

## données préparées
```{r load2}
data_env_topo_clim = readRDS("data_env_topo_clim.rds")
# BIO1 = Annual Mean Temperature
# BIO2 = Mean Diurnal Range (Mean of monthly (max temp - min temp))
# BIO3 = Isothermality (BIO2/BIO7) (×100)
# BIO4 = Temperature Seasonality (standard deviation ×100)
# BIO5 = Max Temperature of Warmest Month
# BIO6 = Min Temperature of Coldest Month
# BIO12 = Annual Precipitation
# BIO15 = Precipitation Seasonality (Coefficient of Variation)
# gdd0 = heat sum of all days above the 0°C temperature accumulated over 1 year
# gsl = Length of the growing season (TREELIM)
# gsp = precipitation sum accumulated on all days during the growing season )TREELIM)
# gst = Mean temperature of all growing season days (TREELIM)
# ngd0 = Number of days at which tas > 0°C
# npp = Calculated based on the ‘Miami model’
# scd = Number of days with snowcover (TREELIM)

data_env_topo_clim$obs$response = 1
data_env_topo_clim$bg$response = 0

dataset = rbind(data_env_topo_clim$obs@data[, -c(1:5)], data_env_topo_clim$bg@data)

dataset = na.omit(dataset[,-which(colnames(dataset)%in% c("aspect", "gsl", "gst"))])
```

## Niche


## Spatial modèle
```{r rfmod, fig = TRUE}
library(randomForest)
set.seed(516)
tuneRF(x=dataset[,c(1:(ncol(dataset)-1))],y=as.factor(dataset$response))

rf.ericar<-randomForest(as.factor(response)~.,mtry=4,ntree=1000,data=dataset)

```

```{r rfspat}
rasalti = rast('ras_alti.tif')
library(spatialRF)
dependent.variable.name <- "response"
predictor.variable.names <- colnames(dataset)[-ncol(dataset)]
dataclean = dataset[which(complete.cases(dataset)),]

xy <- data.frame(rbind(
  coordinates(spTransform(data_env_topo_clim$obs , crs(rasalti))),
  coordinates(spTransform(data_env_topo_clim$bg , crs(rasalti)))
   )[which(complete.cases(dataset)),])
colnames(xy) = c("x", "y")
dim(xy)

dist.mat = dist(xy, diag=T, upper=T)

model.non.spatial <- spatialRF::rf(
  data = dataclean,
  dependent.variable.name = dependent.variable.name,
  predictor.variable.names = predictor.variable.names,
  distance.matrix = as.matrix(dist.mat),
  distance.thresholds = c(5000, 10000),
  xy = xy, 
  seed = 126,
  verbose = TRUE
)

```


<!-- ```{r rfspatspat, cache=TRUE} -->
<!-- model.spatial <- spatialRF::rf_spatial( -->
<!--   model = model.non.spatial, -->
<!--   distance.matrix = distance.matrix, -->
<!--   method = "mem.moran.sequential", #default method -->
<!--   verbose = FALSE, -->
<!--   seed = 123 -->
<!-- ) -->

<!-- ``` -->


<!-- ```{r rfspat performance, fig = TRUE, cache = TRUE, echo=FALSE} -->
<!-- comparison <- spatialRF::rf_compare( -->
<!--   models = list( -->
<!--     `Non-spatial` = model.non.spatial, -->
<!--     `Spatial` = model.spatial -->
<!--   ), -->
<!--   xy = xy, -->
<!--   repetitions = 30, -->
<!--   training.fraction = 0.8, -->
<!--   metrics = "r.squared", -->
<!--   seed = 125 -->
<!-- ) -->

<!-- ``` -->


```{r rfspat importance, fig = TRUE}


p1 <- spatialRF::plot_importance(
  model.non.spatial, 
  verbose = FALSE) + 
  ggplot2::ggtitle("Non-spatial model") 

# p2 <- spatialRF::plot_importance(
#   model.spatial,
#   verbose = FALSE) + 
#   ggplot2::ggtitle("Spatial model")
# 
# p1 | p2 

p1
```

```{r rfspat response, fig = TRUE}
spatialRF::plot_response_curves(
  model.spatial,
  variables = NULL,
  quantiles = 0.5,
  ncol = 3
)
```


## Predict
```{r predictorT, fig = TRUE, cache = TRUE}
topo_rast =  rast("_data/predRast_topo.grd")
clim_rast = rast("_data/predRast_clim.grd")
topo_clip = crop(topo_rast, clim_rast)
rast_envtopo <- c(topo_clip, clim_rast)
plot(rast_envtopo)
# rf.predict.ericar <- predict(rast_envtopo, rf.ericar, type = "prob")
# writeRaster(rf.predict.ericar$X1, file = "rf.predict.ericar.tif", overwrite=TRUE)

```

```{r predict, fig = TRUE}
ras = raster("rf.predict.ericar.tif")
plot(ras)
bbox2 <- tmaptools::bb(xlim = c(6.10,  8.00), ylim = c(44, 46.1))
bbox2_proj = st_bbox(st_transform(st_as_sfc(bbox2), crs(ras)))
ras_crop = terra::crop(ras, extent(bbox2_proj))
 
rf.predict.ericar= read_stars("rf.predict.ericar.tif")
mapview::mapviewOptions(na.color = "transparent",
               basemaps="OpenStreetMap",
               viewer.suppress=TRUE,
               trim=TRUE)
mapview(ras_crop) +
  mapview(st_as_sf(sites_points1), legend = FALSE, cex = 5)
```

